Nesterov's First Order Method Accelerations: Unifications, Applications and Numerial Experiments

### 2025-02-10 Notebook LaTEX/Proposal/presentation.tex
>w<, changes

@@

                \item It demonstrates crucial ideas on how prove convergence rate where the evaluation of proximal point method is inexact in the convex settings. 

@@

\section{Content of the draft paper}
    \begin{frame}{Contents of our draft papers}
        \begin{block}{Upcoming contents}
            We explain results from our relaxed, weak accelerated proximal gradient (R-WAPG). 
            \begin{enumerate}
                \item The R-WAPG sequence and the method of R-WAPG. 
                \item The generic convergence of R-WAPG. 
                \item Three equivalent forms of R-WAPG. 
                \item Specialized convergence results with for specific instances of FISTA variants in the literatures. 
                \item A parameter free formulation of R-WAPG and the results of our preliminary numerical experiments. 
                \item Nesterov's idea of strong convexity transfer and the R-WAPG framework. 
            \end{enumerate}
        \end{block}

@@

        \begin{frame}{R-WAPG momentum form}
            \begin{definition}[R-WAPG momentum form]\label{def:r-wapg-momentum-form}
                Given any $y_1 = x_1 \in \RR^n$, and sequences $(\rho_k)_{k \ge 0}, (\alpha_k)_{k\ge 0}$ Definition \ref{def:rwapg-seq}. 
                The algorithm generates iterates $x_{k + 1}, y_{k + 1}$ For $k = 1, 2, \cdots $ by the procedures: 
                {\small
                \begin{tcolorbox}
                    For $k=1, 2,\ldots $
                    \begin{align*}
                        & x_{k + 1} = y_k - L^{-1}\mathcal G_Ly_k, 
                        \\
                        & 
                        y_{k + 1} = 
                        x_{k + 1} + 
                        \frac{\rho_k\alpha_k(1 - \alpha_k)}{\rho_k\alpha_k^2 + \alpha_{k + 1}}(x_{k + 1} - x_k). 
                    \end{align*}    
                \end{tcolorbox}
                }
                In the special case where $\mu = 0$, the momentum term can be represented without parameter $\rho_k$: 
                $$
                    (\forall k \ge 1)\quad \frac{\rho_k\alpha_k(1 - \alpha_k)}{\rho_k\alpha_k^2 + \alpha_{k + 1}} 
                    = \alpha_{k + 1}(\alpha_k^{-1} - 1). 
                $$
            \end{definition}
        \end{frame}

    \subsection{Unified Convergence claim with relaxed Nesterov's sequence}
        \begin{frame}{Summary of our results}
            With the equivalent representations and the convergence claim for relaxed sequence $(\alpha_k)_{k \ge 0}$ of the R-WAPG, we are able to unifies: 
            \begin{enumerate}
                \item Several Euclidean variants of the FISTA algorithm. 
                \item Nontraditional choices of momentum sequences. 
            \end{enumerate}
            The table below summarizes our major results. \begin{table}[H]
            {\tiny
            \begin{tabular}{|l|l|l|l|}
            \hline
                Algorithm & $\mu$ & $\alpha_k, \rho_k$ & $F(x_k) - F^* \le \mathcal O(\cdot)$ 
            \\ \hline
                Definition \ref{def:wapg} & 
                $\mu \ge 0$ &
                $\alpha_k \in(\mu/L, 1)$, $\rho_k > 0$ & 
                \begin{tabular}{l}
                    $\prod_{i = 0}^{k-1} \max(1, \rho_i)(1 - \alpha_{i + 1})$
                    \\
                    (Proposition \ref{prop:wagp-convergence})
                \end{tabular}
            \\ \hline
                FISTA \cite{chambolle_convergence_2015} &  
                $\mu = 0$  &
                $ 0< \alpha_k^{-2} \le \alpha_{k + 1}^{-1} - \alpha_{k + 1}^{-2}$, $\rho_k \ge 1$ &
                \begin{tabular}{l}
                    $\alpha_k^{2}$ 
                    % \\ 
                    % (Theorem \ref{thm:r-wapg-on-cham-doss})    
                \end{tabular}
            \\ \hline
                \hspace{-1em}\begin{tabular}{l}
                    V-FISTA \\ (10.7.7) \cite{beck_first-order_2017} 
                \end{tabular} &
                $\mu > 0$ &
                $\alpha_k = \sqrt{\mu/L}$, $\rho_k = 1$ & 
                \begin{tabular}{l}
                    $(1 - \sqrt{\mu/L})^k$, 
                    % \\
                    % (Theorem \ref{thm:fixed-momentum-fista}, remark)
                \end{tabular}
            \\ \hline
                Definition \ref{def:wapg} &  
                $\mu > 0$ &
                $\alpha_k = \alpha \in (\mu/L, 1)$, $\rho_k = \rho > 0$ &  
                \begin{tabular}{l}
                    $\max(1 - \alpha, 1 - \mu/(\alpha L))^{k}$
                    % \\
                    % (Theorem \ref{thm:fixed-momentum-fista})
                \end{tabular}
            \\ \hline
            \end{tabular}
            }
            \end{table}
            These results are consistent of literatures. 
            To the best of our knowledge, the last variant is, and we have the convergence claim for it using R-WAPG. 
        \end{frame}
    \subsection{A parameter free formulation of R-WAPG}
        \begin{frame}{Free R-WAPG}
            We proposed the following implementation of R-WAPG which doesn't require parameters $\mu, L$ in advance.

            \begin{algorithm}[H]
            {\scriptsize
                \begin{algorithmic}
                \STATE{\textbf{Input: } $f, g, x_0, L > \mu \ge 0, \in \RR^n, N \in \N$}
                \STATE{\textbf{Initialize: }$y_0 := x_0;L := 1; \mu := 1/2; \alpha_0 = 1;$}
                \STATE{\textbf{Compute: } $f(y_k)$; }
                \FOR{$k = 0, 1, 2, \cdots, N$}
                    \STATE{\textbf{Compute: }$\nabla f(y_k); x^+:= [I + L^{-1}\partial g](y_k - L^{-1}\nabla f(y_k))$;}
                    \WHILE{$L/2\Vert x^+ - y\Vert^2 < D_f(x^+, y)$}
                        \STATE{$L:= 2L$;}
                        \STATE{$x^+ = [I + L^{-1}\partial g](y_k - L^{-1}\nabla f(y_k))$; }
                    \ENDWHILE
                    \STATE{$x_{k + 1} := x^+$;}
                    \STATE{$\alpha_{k + 1} := (1/2)\left(\mu/L - \alpha_{k}^2 + \sqrt{(\mu/L - \alpha_{k}^2)^2 + 4\alpha_{k}^2}\right)$;}
                    \STATE{$\theta_{k + 1} := \alpha_k(1 - \alpha_k)/(\alpha_k^2 + \alpha_{k + 1})$;}
                    \STATE{$y_{k + 1}:= x_{k + 1} + \theta_{k + 1}(x_{k + 1} - x_k)$; }
                    \STATE{\textbf{Compute: } $f(y_{k + 1})$}
                    \STATE{$\mu := (1/2)(2D_f(y_{k + 1}, y_{k})/\Vert y_{k + 1} - y_k\Vert^2) + (1/2)\mu$;}
                \ENDFOR
                \end{algorithmic}
                \caption{Free R-WAPG}
                \label{alg:free-rwapg}
            }
            \end{algorithm}
        \end{frame}

@@

@@

                \item Same initial guess shared by FR-WAPG from us, and  M-FISTA, V-FISTA in (10.7.7, 10.7.6) by Beck \cite{beck_first-order_2017}. It's repeated for 30 different random initial guesses. 

@@

                \item $\mu, L$ were given in prior to produce diagonal matrix $A = \text{diag}(0, \mu + (L-\mu)(N - 1)^{-1}, \mu + 2(L-\mu)(N - 1)^{-1}, \cdots, \mu + (N - 2)(L - \mu)^{-1}, L)$, but M-FISTA, FR-WAPG were not fed these parameters. 

@@

@@

                \item Structure of Catalyst accelerations and inner/out loop complexity results from Lin et al's first Catalyst paper \cite{lin_universal_2015}. 
                \item Catalyst acceleration with relative error crition, Lin et al's second paper \cite{lin_catalyst_2018}.

@@

                            \mathcal M_F^{\kappa^{-1}}(z; x) - 
                            \mathcal M^*_{F, \kappa^{-1}}(z; x)

@@

        \begin{frame}{Outer loop complexity, strong convex case}

@@

        \begin{frame}{Outer loop complexity, convex but not strongly convex case}

@@

            The followings are Proposition 3.2, 3.4 from Lin et al. \cite{lin_universal_2015}. 
            \begin{proposition}[Inner loop complexity strongly convex]\label{prop:inner-loop-complexity-s-cnvx}

@@

                Here $\widetilde{\mathcal O}$ hides logarithmic complexity in $\mu, \kappa$ and other constants. 

@@

            \begin{proposition}[Inner loop, convex but not strongly convex]\label{prop:inner-loop-complexity-cnvx}

@@

            We count $m$, the number of iteration experienced by $\mathbb M$ for the $k$ th iteration of $\mathbb A$. 

@@

                \item If $\mu = 0$, using Proposition \ref{prop:inner-loop-complexity-cnvx}, Theorem \ref{thm:erro-seq-outer-cnvx} it has 

@@

                \item $N_{\mathbb N}\le \widetilde{O}\left((\tau_{\mathbb M}\sqrt{q})^{-1}\log(\epsilon^{-1})\right)$ if $\mu > 0$. 

@@

                \item $\delta_k$ doesn't requite $F^*$. 

@@

                Extend linear convergence of accelerated gradient for variance reduced incremental methods such as the Catalyst Framework into a weaker settings. 

@@

        \end{frame}
        \begin{frame}{Our question on Catalyst Acceleration}
            \begin{block}{A key detail about Catalyst}
                The smoothness assumption is baked into Assumption \ref{ass:linear-convergence-inner-loop}, where it states that the $\mathbb M$ needs to have linear convergence rate. 
            \end{block}
            All VRMs method in the literature requires Lipschitz smoothness and strong convexity to achieve linear convergence rate. 
            No linear convergent first order algorithm exists for all strong convex function by Theorem 2.1.5 in Nesterov's book \cite{nesterov_lectures_2018}. 
            Hence, here is our question: 
            \begin{alertblock}{Our question}
                Solving inexact proximal point doesn't necessarily need smoothness assumption on $F$ so what if we Assumption \ref{ass:linear-convergence-inner-loop} is false? 
            \end{alertblock}
            Next slide we explain why the answer is not trivial and why it's interesting. 
        \end{frame}
        \begin{frame}{The answer to our question is nontrivial}
            Without smoothness assumption on $F$, the following considerations apply. 
            \begin{enumerate}
                \item $\mathbb A$ optimizes inexact Moreau Envelope $\mathcal M^*_{\kappa^{-1}, F}(y)$ which is smooth and strongly convex if $F$ is convex hence the $\mathcal O(1/\sqrt{k})$ lower bound of Theorem 3.2.1 \cite{nesterov_lectures_2018} complexity for all convex first order optimization doesn't apply for the total complexity.  
                \item Instead, Theorem 2.1.7 \cite{nesterov_lectures_2018} applies and the lower complexity bound $\mathbb A$ is $\mathcal O(1/k^2)$. 
                \item Linear convergence of $\mathbb M$ Theorem 2.1.17 fails instead Theorem 3.2.5 applies and the lower bound is $\mathcal O(1/k)$. 
            \end{enumerate}
            We note that the question has practical importance because the method of subgradient descent with Polyak stepsizes are optimal for Smooth, strongly convex, and smooth plus strongly convex function on a compact domain. See Hazan and Kakade\cite{hazan_revisiting_2022}. 
        \end{frame}

>>>>

        Part I: Introduction. 

@@

                \item It demonstrates crucial ideas on convergence analysis involving nexact evaluation of proximal point method under the the convex settings. 

@@

\section{Contents of the draft paper}
    \begin{frame}{Contents of our draft papers}
        Part II: Contents of our current draft paper. 
        \begin{block}{Upcoming contents}
            We explain results from our relaxed, weak accelerated proximal gradient (R-WAPG). 
            \begin{enumerate}
                \item The R-WAPG sequence and the method of R-WAPG. 
                \item A generic convergence theorem of R-WAPG. 
                \item Three equivalent forms of R-WAPG. 
                \item Specialized convergence results with for specific instances of FISTA variants in the literatures. 
                \item A parameter free formulation of R-WAPG and the results of our preliminary numerical experiments. 
                \item Nesterov's idea of strong convexity transfer and the R-WAPG framework. 
            \end{enumerate}
        \end{block}

@@

        \begin{frame}{R-WAPG momentum form}
            \begin{definition}[R-WAPG momentum form]\label{def:r-wapg-momentum-form}
                Given any $y_1 = x_1 \in \RR^n$, and sequences $(\rho_k)_{k \ge 0}, (\alpha_k)_{k\ge 0}$ Definition \ref{def:rwapg-seq}. 
                The algorithm generates iterates $x_{k + 1}, y_{k + 1}$ For $k = 1, 2, \ldots $ by the procedures: 
                {\small
                \begin{tcolorbox}
                    For $k=1, 2,\ldots $
                    \begin{align*}
                        & x_{k + 1} = y_k - L^{-1}\mathcal G_Ly_k, 
                        \\
                        & 
                        y_{k + 1} = 
                        x_{k + 1} + 
                        \frac{\rho_k\alpha_k(1 - \alpha_k)}{\rho_k\alpha_k^2 + \alpha_{k + 1}}(x_{k + 1} - x_k). 
                    \end{align*}    
                \end{tcolorbox}
                }
                In the special case where $\mu = 0$, the momentum term can be represented without parameter $\rho_k$: 
                $$
                    (\forall k \ge 1)\quad \frac{\rho_k\alpha_k(1 - \alpha_k)}{\rho_k\alpha_k^2 + \alpha_{k + 1}} 
                    = \alpha_{k + 1}(\alpha_k^{-1} - 1). 
                $$
            \end{definition}
        \end{frame}

    \subsection{Unified Convergence claim with relaxed Nesterov's sequence}
        \begin{frame}{Summary of our results}
            With the equivalent representations and the convergence claim for relaxed sequence $(\alpha_k)_{k \ge 0}$ of the R-WAPG, we are able to unify the convergence rate of: 
            \begin{enumerate}
                \item Three Euclidean variants of the FISTA algorithm. 
                \item Constant momentum APG for strongly convex objective.  
            \end{enumerate}
            The table below summarizes our major results. \begin{table}[H]
            {\tiny
            \begin{tabular}{|l|l|l|l|}
            \hline
                Algorithm & $\mu$ & $\alpha_k, \rho_k$ & $F(x_k) - F^* \le \mathcal O(\cdot)$ 
            \\ \hline
                Definition \ref{def:wapg} & 
                $\mu \ge 0$ &
                $\alpha_k \in(\mu/L, 1)$, $\rho_k > 0$ & 
                \begin{tabular}{l}
                    $\prod_{i = 0}^{k-1} \max(1, \rho_i)(1 - \alpha_{i + 1})$
                    \\
                    (Proposition \ref{prop:wagp-convergence})
                \end{tabular}
            \\ \hline
                FISTA \cite{chambolle_convergence_2015} &  
                $\mu = 0$  &
                $ 0< \alpha_k^{-2} \le \alpha_{k + 1}^{-1} - \alpha_{k + 1}^{-2}$, $\rho_k \ge 1$ &
                \begin{tabular}{l}
                    $\alpha_k^{2}$ 
                    % \\ 
                    % (Theorem \ref{thm:r-wapg-on-cham-doss})    
                \end{tabular}
            \\ \hline
                \hspace{-1em}\begin{tabular}{l}
                    V-FISTA \\ (10.7.7) \cite{beck_first-order_2017} 
                \end{tabular} &
                $\mu > 0$ &
                $\alpha_k = \sqrt{\mu/L}$, $\rho_k = 1$ & 
                \begin{tabular}{l}
                    $(1 - \sqrt{\mu/L})^k$, 
                    % \\
                    % (Theorem \ref{thm:fixed-momentum-fista}, remark)
                \end{tabular}
            \\ \hline
                Definition \ref{def:wapg} &  
                $\mu > 0$ &
                $\alpha_k = \alpha \in (\mu/L, 1)$, $\rho_k = \rho > 0$ &  
                \begin{tabular}{l}
                    $\max(1 - \alpha, 1 - \mu/(\alpha L))^{k}$
                    % \\
                    % (Theorem \ref{thm:fixed-momentum-fista})
                \end{tabular}
            \\ \hline
            \end{tabular}
            }
            \end{table}
            These results are consistent of literatures. 
            To the best of our knowledge, the last variant is, and we have the convergence claim for it using R-WAPG. 
        \end{frame}
    \subsection{A parameter free formulation of R-WAPG}
        \begin{frame}{Free R-WAPG}
            We proposed the following implementation of R-WAPG which doesn't require parameters $\mu, L$ in advance.

            \begin{algorithm}[H]
            {\scriptsize
                \begin{algorithmic}
                \STATE{\textbf{Input: } $f, g, x_0, L > \mu \ge 0, \in \RR^n, N \in \N$}
                \STATE{\textbf{Initialize: }$y_0 := x_0;L := 1; \mu := 1/2; \alpha_0 = 1;$}
                \STATE{\textbf{Compute: } $f(y_k)$; }
                \FOR{$k = 0, 1, 2, \cdots, N$}
                    \STATE{\textbf{Compute: }$\nabla f(y_k); x^+:= [I + L^{-1}\partial g](y_k - L^{-1}\nabla f(y_k))$;}
                    \WHILE{$L/2\Vert x^+ - y_k\Vert^2 < D_f(x^+, y_k)$}
                        \STATE{$L:= 2L$;}
                        \STATE{$x^+ = [I + L^{-1}\partial g](y_k - L^{-1}\nabla f(y_k))$; }
                    \ENDWHILE
                    \STATE{$x_{k + 1} := x^+$;}
                    \STATE{$\alpha_{k + 1} := (1/2)\left(\mu/L - \alpha_{k}^2 + \sqrt{(\mu/L - \alpha_{k}^2)^2 + 4\alpha_{k}^2}\right)$;}
                    \STATE{$\theta_{k + 1} := \alpha_k(1 - \alpha_k)/(\alpha_k^2 + \alpha_{k + 1})$;}
                    \STATE{$y_{k + 1}:= x_{k + 1} + \theta_{k + 1}(x_{k + 1} - x_k)$; }
                    \STATE{\textbf{Compute: } $f(y_{k + 1})$}
                    \STATE{$\mu := (1/2)(2D_f(y_{k + 1}, y_{k})/\Vert y_{k + 1} - y_k\Vert^2) + (1/2)\mu$;}
                \ENDFOR
                \end{algorithmic}
                \caption{Free R-WAPG}
                \label{alg:free-rwapg}
            }
            \end{algorithm}
        \end{frame}

@@

                    \NOG
                \text{ with } 

@@

                \item Same initial guess $x_0$ shared by FR-WAPG,  M-FISTA and V-FISTA in (10.7.7, 10.7.6) by Beck \cite{beck_first-order_2017} repeated with 30 different random initial guesses. 

@@

                \item $\mu, L$ were given in prior to produce diagonal matrix 
                    A = \text{diag}
                    \left(0, 
                        \mu + \frac{L - \mu}{N - 1},
                        \mu + 2\frac{L - \mu}{N - 1}, 
                        \ldots, 
                        \mu + \frac{(N - 2)(L - \mu)}{N - 1}, 
                        L
                M-FISTA, FR-WAPG were not fed these parameters. 

@@

        Part III: Catalyst Acceleration and future works. 

@@

                \item Structure of Catalyst accelerations and inner/out loop complexity results from Lin et al.'s first Catalyst paper \cite{lin_universal_2015}. 
                \item Catalyst acceleration with relative error criterion, Lin et al.'s second paper \cite{lin_catalyst_2018}.

@@

                            \mathcal M_F^{1/\kappa}(z; x) - 
                            \mathcal M^*_{F, 1/\kappa}(z; x)

@@

        \begin{frame}{Outer loop complexity, $\mu > 0$}

@@

        \begin{frame}{Outer loop complexity, $\mu \ge 0$}

@@

            The followings are Proposition 3.2, 3.4 from Lin et al. 
            \cite{lin_universal_2015}. 
            \begin{proposition}[Inner loop complexity $\mu > 0$]
            \label{prop:inner-loop-complexity-s-cnvx}

@@

                Here $\widetilde{\mathcal O}$ hides logarithmic complexity, $\mu, \kappa$ and other constants. 

@@

            \begin{proposition}[Inner loop complexity $\mu \ge 0$]
            \label{prop:inner-loop-complexity-cnvx}

@@

            We count $m$, the number of iteration experienced by $\mathbb M$ up to the $k$ iteration of $\mathbb A$. 

@@

                \item If $\mu = 0$ so $F$ is not necessarily strongly convex, using Proposition \ref{prop:inner-loop-complexity-cnvx}, Theorem \ref{thm:erro-seq-outer-cnvx} it has 

@@

                \item $N_{\mathbb N}\le \widetilde{\mathcal O}\left((\tau_{\mathbb M}\sqrt{q})^{-1}\log(\epsilon^{-1})\right)$ if $\mu > 0$. 

@@

                \item Relative error sequence $\delta_k$ doesn't requite $F^*$. 

@@

                Extend linear convergence of accelerated gradient for variance reduced incremental methods such as the Catalyst Framework into a weaker setting. 

@@

        \end{frame}
        \begin{frame}{Our question on Catalyst Acceleration}
            \begin{block}{A key detail about Catalyst}
                The smoothness assumption is baked into Assumption \ref{ass:linear-convergence-inner-loop}, where it states that the $\mathbb M$ needs to have linear convergence rate. 
            \end{block}
            All VRMs method in the literature requires Lipschitz smoothness and strong convexity to achieve linear convergence rate. 
            No linear convergent first order algorithm exists for all strong convex function by Theorem 2.1.5 in Nesterov's book \cite{nesterov_lectures_2018}. 
            Hence, here is our question: 
            \begin{alertblock}{Our question}
                Solving inexact proximal point doesn't necessarily need smoothness assumption on $F$ so what if we Assumption \ref{ass:linear-convergence-inner-loop} is false? 
            \end{alertblock}
            Next slide we explain why the answer is not trivial and why it's interesting. 
        \end{frame}
        \begin{frame}{The answer to our question is nontrivial}
            Without smoothness assumption on $F$, the following considerations apply. 
            \begin{enumerate}
                \item $\mathbb A$ optimizes inexact Moreau Envelope $\mathcal M^*_{\kappa^{-1}, F}(y)$ which is smooth and strongly convex if $F$ is convex hence $\mathcal O(1/\sqrt{k})$ from Theorem 3.2.1 \cite{nesterov_lectures_2018} doesn't apply for the total complexity.  
                \item Theorem 2.1.7 \cite{nesterov_lectures_2018} doesn't apply because the lower complexity bound $\mathbb A$ is $\mathcal O(1/k^2)$ assume exact gradient evaluations. 
                \item Theorem 2.1.17 \cite{nesterov_lectures_2018} (Linear rate lower bound for strong convexity and smoothness) doesn't apply on $\mathbb M$. 
                Instead Theorem 3.2.5 applies and the lower bound is $\mathcal O(1/k)$. 
            \end{enumerate}
            Subgradient descent with Polyak step size reaches the lower bound of worst case of Strongly convex, strongly convex and smooth.
            See Hazan et al. \cite{hazan_revisiting_2022}.
